---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file and knit. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 8,
  fig.height = 3.4,
  dpi = 120,
  out.width = "100%"
)
```

# msamodel

<!-- badges: start -->
[![R-CMD-check](https://github.com/jechave/msamodel/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jechave/msamodel/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Across a family of homologous enzymes, some residues change structure a lot and
others almost not at all. `msamodel` predicts this divergence profile from a single
structure and its active site, using the Mutation-Stability-Activity model, and
splits it into the contributions of mutation, stability selection, and activity
selection.

## Installation

```r
# install.packages("remotes")
remotes::install_github("jechave/msamodel", dependencies = TRUE)
```

The elastic network model comes from [penm](https://github.com/jechave/penm),
installed alongside. Neither package is on CRAN.

## Example

Mutate every site, fit the model to an observed profile, predict:

```{r fit, message = FALSE}
library(msamodel)
library(penm)

ex     <- function(f) system.file("extdata", f, package = "msamodel")
active <- read.csv(ex("1d6o_A_active_site.csv"))
obs    <- read.csv(ex("1d6o_A_lrmsd_obs_site.csv"))

wt  <- penm::set_enm(bio3d::read.pdb(ex("1d6o_A.pdb")), node = "ca",
                     model = "ming_wall", d_max = 10.5, frustrated = FALSE)
spm <- generate_spm(wt, n_mutations = 10,
                    pdb_site_active = active$pdb_site, ensemble = 1L)

fit  <- fit_lrmsd_msa_site(spm, obs$pdb_site, obs$lrmsd_obs)
pred <- predict_profiles(fit, spm, metric = "nlrmsd")$site
```

```{r profile, echo = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

pred <- pred |>
  inner_join(obs, by = "pdb_site") |>
  mutate(observed = lrmsd_obs - mean(lrmsd_obs),
         lower = nlrmsd_msa - 1.96 * nlrmsd_msa_se,
         upper = nlrmsd_msa + 1.96 * nlrmsd_msa_se)

ggplot(pred, aes(pdb_site)) +
  geom_vline(xintercept = active$pdb_site, colour = "grey60",
             linetype = 2, linewidth = 0.35) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#C41E3A", alpha = 0.18) +
  geom_line(aes(y = nlrmsd_msa), colour = "#C41E3A", linewidth = 0.6) +
  geom_point(aes(y = observed), colour = "grey30", size = 0.9, alpha = 0.85) +
  labs(x = "Residue", y = "Structural divergence") +
  theme_minimal(base_size = 11)
```

Observed divergence (points), fitted model with 95% band, active-site residues
dashed. The fit has two parameters — selection strengths on stability
(`r sprintf("%.2f", fit$a1)`) and on activity (`r sprintf("%.0f", fit$a2)`) — and
accounts for `r sprintf("%.0f%%", 100 * fit$gof$D2)` of the variance in the
observed profile.

The same fit splits the profile into its three constituent parts:

```{r decomp-call}
dec <- predict_decomposition(fit, spm, metric = "nlrmsd")$site
```

```{r decomposition, echo = FALSE, message = FALSE, warning = FALSE}
dec <- dec |>
  select(pdb_site, nphi_mut, nphi_stab, nphi_act) |>
  pivot_longer(-pdb_site, names_to = "component", values_to = "phi") |>
  mutate(component = factor(component,
                            levels = c("nphi_mut", "nphi_stab", "nphi_act"),
                            labels = c("Mutation", "Stability", "Activity")))

ggplot(dec, aes(pdb_site, phi, colour = component)) +
  geom_vline(xintercept = active$pdb_site, colour = "grey60",
             linetype = 2, linewidth = 0.35) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_line(linewidth = 0.55) +
  scale_colour_manual(values = c(Mutation = "#FF8C00", Stability = "#0000CD",
                                 Activity = "#C41E3A")) +
  labs(x = "Residue", y = "Contribution", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

The three curves add up to the profile above. For this enzyme activity selection
does most of the work and stability almost none; other families differ.

## Interface

`generate_spm()` mutates every site and records the structural response. It is the
slow step, and everything else works from its output.

`calculate_profiles()` and `calculate_decomposition()` evaluate the model at
selection strengths given as arguments — for exploring how each constraint shapes
the profile. `predict_profiles()` and `predict_decomposition()` evaluate it at a
fit obtained from data, with error bands. Fits come from `fit_lrmsd_msa_site()`
and `fit_lrmsd_msa_mode()`.

All four return a `$site` table and a `$mode` table: divergence per residue, and
per normal mode of the structure.

Inputs are a PDB file, active-site residue numbers, and an observed profile to fit
against. Measuring that profile — superimposing homologs over an alignment — is
outside this package.

## Documentation

```r
?msamodel                        # API, indexed by workflow step
vignette("msamodel")             # both axes, end to end
vignette("msamodel-explore")     # profiles at parameters you choose, both axes
vignette("msamodel-fit")         # fitting to data, with confidence bands
```

## Reference

Echave J, Carpentier M (2026). Why structural divergence varies among residues in
enzyme evolution: contributions of mutation, stability, and activity constraints.
*Molecular Biology and Evolution* 43(7), msag162.
<https://doi.org/10.1093/molbev/msag162>

## License

MIT © Julian Echave
