---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file and knit. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 8,
  fig.height = 3.4,
  dpi = 120,
  out.width = "100%"
)
```

# msamodel

<!-- badges: start -->
[![R-CMD-check](https://github.com/jechave/msamodel/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jechave/msamodel/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `msamodel` package implements the Mutation-Stability-Activity (MSA) model of
protein structural evolution. In the MSA model, mutations perturb the structure and
become fixed or are lost depending on their effects on stability and activity,
weighted by stability and activity selection strengths, respectively. Given a single
structure and its active-site residues, the model predicts the resulting structural
divergence profiles, per residue and per normal mode. The two selection strengths can
be set to chosen values, to explore how each constraint shapes the profiles, or
estimated from observed profiles by maximum likelihood. In addition, predicted
profiles can be analysed in terms of nested "what-if" scenarios in which one or both
selection constraints are dropped, and can be decomposed as a sum of mutation,
stability, and activity contributions.

## Installation

```r
# install.packages("remotes")
remotes::install_github("jechave/msamodel", dependencies = TRUE)
```

The elastic network model comes from [penm](https://github.com/jechave/penm),
installed alongside. Neither package is on CRAN.

## Example

Mutate every site, fit the model to an observed profile, predict:

```{r fit, message = FALSE}
library(msamodel)
library(penm)

ex     <- function(f) system.file("extdata", f, package = "msamodel")
active <- read.csv(ex("1d6o_A_active_site.csv"))
obs    <- read.csv(ex("1d6o_A_lrmsd_obs_site.csv"))

wt  <- penm::set_enm(bio3d::read.pdb(ex("1d6o_A.pdb")), node = "ca",
                     model = "ming_wall", d_max = 10.5, frustrated = FALSE)
spm <- generate_spm(wt, n_mutations = 10,
                    pdb_site_active = active$pdb_site, ensemble = 1L)

fit  <- fit_lrmsd_msa_site(spm, obs$pdb_site, obs$lrmsd_obs)
pred <- predict_profiles(fit, spm, metric = "nlrmsd")$site
```

```{r profile, echo = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

pred <- pred |>
  inner_join(obs, by = "pdb_site") |>
  mutate(observed = lrmsd_obs - mean(lrmsd_obs),
         lower = nlrmsd_msa - 1.96 * nlrmsd_msa_se,
         upper = nlrmsd_msa + 1.96 * nlrmsd_msa_se)

ggplot(pred, aes(pdb_site)) +
  geom_vline(xintercept = active$pdb_site, colour = "grey60",
             linetype = 2, linewidth = 0.35) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#C41E3A", alpha = 0.18) +
  geom_line(aes(y = nlrmsd_msa), colour = "#C41E3A", linewidth = 0.6) +
  geom_point(aes(y = observed), colour = "grey30", size = 0.9, alpha = 0.85) +
  labs(x = "Residue", y = "Structural divergence") +
  theme_minimal(base_size = 11)
```

Observed divergence (points), fitted model with 95% band, active-site residues
dashed. The two fitted selection strengths are
`r sprintf("%.2f", fit$a1)` on stability and `r sprintf("%.0f", fit$a2)` on
activity, accounting for `r sprintf("%.0f%%", 100 * fit$gof$D2)` of the variance
in the observed profile.

The same fit decomposes the profile into its three contributions:

```{r decomp-call}
dec <- predict_decomposition(fit, spm, metric = "nlrmsd")$site
```

```{r decomposition, echo = FALSE, message = FALSE, warning = FALSE}
dec <- dec |>
  select(pdb_site, nphi_mut, nphi_stab, nphi_act) |>
  pivot_longer(-pdb_site, names_to = "component", values_to = "phi") |>
  mutate(component = factor(component,
                            levels = c("nphi_mut", "nphi_stab", "nphi_act"),
                            labels = c("Mutation", "Stability", "Activity")))

ggplot(dec, aes(pdb_site, phi, colour = component)) +
  geom_vline(xintercept = active$pdb_site, colour = "grey60",
             linetype = 2, linewidth = 0.35) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_line(linewidth = 0.55) +
  scale_colour_manual(values = c(Mutation = "#FF8C00", Stability = "#0000CD",
                                 Activity = "#C41E3A")) +
  labs(x = "Residue", y = "Contribution", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

The three contributions add up to the profile above. Their balance varies from
family to family.

## Interface

`generate_spm()` mutates every site and records the structural response.
Everything else works from its output.

`calculate_profiles()` and `calculate_decomposition()` evaluate the model at
selection strengths set to chosen values, to explore how each constraint shapes the
profiles. `predict_profiles()` and `predict_decomposition()` evaluate it at a fit
obtained from data, with error bands. Fits come from `fit_lrmsd_msa_site()` and
`fit_lrmsd_msa_mode()`.

All four return a `$site` table and a `$mode` table, holding the profiles on the
residue axis and on the normal-mode axis.

Inputs are a PDB file, active-site residue numbers, and an observed profile to fit
against. Measuring that profile, by superimposing homologs over an alignment, is
outside this package.

## Documentation

```r
?msamodel                        # API, indexed by workflow step
vignette("msamodel")             # both axes, end to end
vignette("msamodel-explore")     # profiles at chosen parameters, both axes
vignette("msamodel-fit")         # fitting to data, with confidence bands
```

## Reference

Echave J, Carpentier M (2026). Why structural divergence varies among residues in
enzyme evolution: contributions of mutation, stability, and activity constraints.
*Molecular Biology and Evolution* 43(7), msag162.
<https://doi.org/10.1093/molbev/msag162>

## License

MIT © Julian Echave
