---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file and knit. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 8,
  fig.height = 3.4,
  dpi = 120,
  out.width = "100%"
)
```

# msamodel

<!-- badges: start -->
[![R-CMD-check](https://github.com/jechave/msamodel/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jechave/msamodel/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `msamodel` package implements the Mutation-Stability-Activity (MSA) model of
protein structural evolution. In the MSA model, mutations perturb the structure and
become fixed or are lost depending on their effects on stability and activity,
weighted by stability and activity selection strengths, respectively. Given a single
structure and its active-site residues, the model predicts the resulting structural
divergence profiles, per residue and per normal mode. The two selection strengths can
be set to chosen values, to explore how each constraint shapes the profiles, or
estimated from observed profiles by maximum likelihood. In addition, predicted
profiles can be analysed in terms of nested "what-if" scenarios in which one or both
selection constraints are dropped, and can be decomposed as a sum of mutation,
stability, and activity contributions.

## Installation

Run this at an R prompt, not in the shell:

```r
# install.packages("remotes")
remotes::install_github("jechave/msamodel", dependencies = TRUE)
```

The elastic network model comes from [penm](https://github.com/jechave/penm),
installed alongside. Neither package is on CRAN.

Every code block below is R as well, to be run in the same session after
`library(msamodel)`.

## Interface

The inputs are a PDB file and a vector of active-site residue numbers. Fitting also
takes an observed divergence profile, as two vectors: the residue numbers, and the
divergence at each. Measuring that profile, by superimposing homologs over an
alignment, is outside this package.

| Function | Does | Returns |
|---|---|---|
| `generate_spm()` | mutates every site and records the structural response | the scan every other function reads |
| `fit_lrmsd_msa_site()` | estimates the two selection strengths by fitting an observed site profile | the fit, with standard errors and `$gof` |
| `fit_lrmsd_msa_mode()` | estimates them by fitting an observed mode profile | the fit, with standard errors and `$gof` |
| `calculate_profiles()` | evaluates the model at selection strengths given as arguments | the profiles |
| `calculate_decomposition()` | decomposes the model at selection strengths given as arguments | the nested models and the three contributions |
| `predict_profiles()` | evaluates the model at a fit | the profiles, with standard errors |
| `predict_decomposition()` | decomposes the model at a fit | the nested models and the three contributions, with standard errors |

The `calculate_*` pair explores how each constraint shapes the profiles; the
`predict_*` pair evaluates the model where the data puts it.

Every profile comes in two representations, and the `calculate_*` and `predict_*`
functions return both: the same divergence written in two bases. The `$site` tibble
gives it residue by residue, the `$mode` tibble gives it per normal mode. Their
`metric` argument chooses the quantity: `"lrmsd"` keeps the profile's overall level, `"nlrmsd"` centres it on its
mean, which is the form a fit is made on.

## Documentation

`?msamodel` is the package page: the workflow in order, with a link to each
function.

Three vignettes, each a worked example on the shipped protein:

```r
vignette("msamodel")             # the whole workflow, scan to decomposition
vignette("msamodel-explore")     # the model at chosen selection strengths
vignette("msamodel-fit")         # estimating those strengths from data
```

## Example

`penm::set_enm()` builds the elastic network from the structure, and
`generate_spm()` mutates every site and records how the structure responds.

```{r scan, message = FALSE}
library(msamodel)
library(penm)

ex     <- function(f) system.file("extdata", f, package = "msamodel")
active <- read.csv(ex("1d6o_A_active_site.csv"))
obs    <- read.csv(ex("1d6o_A_lrmsd_obs_site.csv"))

wt  <- penm::set_enm(bio3d::read.pdb(ex("1d6o_A.pdb")), node = "ca",
                     model = "ming_wall", d_max = 10.5, frustrated = FALSE)
spm <- generate_spm(wt, n_mutations = 10,
                    pdb_site_active = active$pdb_site, ensemble = 1L)
```

`fit_lrmsd_msa_site()` estimates the two selection strengths from an observed
profile, given as two vectors: the residue numbers and the divergence at each. The
fit carries the estimates, their standard errors, and a goodness-of-fit summary in
`$gof`.

```{r fit}
fit <- fit_lrmsd_msa_site(spm, obs$pdb_site, obs$lrmsd_obs)

c(a1 = fit$a1, a2 = fit$a2)
fit$gof
```

`predict_profiles()` evaluates the model at that fit. It returns a `$site` and a
`$mode` tibble, each holding the profile and its standard error in one
representation.

```{r profiles-call}
pred <- predict_profiles(fit, spm, metric = "nlrmsd")

pred$site
pred$mode
```

Plotting both: the observed profile exists only per residue, so the points appear
in the left panel alone.

```{r profiles, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3.2}
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

n_modes_shown <- 50

theme_readme <- theme_minimal(base_size = 10) +
  theme(legend.position = "top", legend.title = element_blank(),
        plot.title = element_text(size = 10, face = "bold"))

active_lines <- geom_vline(xintercept = active$pdb_site, colour = "grey60",
                           linetype = 2, linewidth = 0.3)

crop_modes <- function(d) filter(d, mode <= n_modes_shown)

site <- pred$site |>
  inner_join(obs, by = "pdb_site") |>
  mutate(observed = lrmsd_obs - mean(lrmsd_obs))

p_site <- ggplot(site, aes(pdb_site)) +
  active_lines +
  geom_ribbon(aes(ymin = nlrmsd_msa - 1.96 * nlrmsd_msa_se,
                  ymax = nlrmsd_msa + 1.96 * nlrmsd_msa_se),
              fill = "#C41E3A", alpha = 0.18) +
  geom_line(aes(y = nlrmsd_msa), colour = "#C41E3A", linewidth = 0.6) +
  geom_point(aes(y = observed), colour = "grey25", size = 0.8, alpha = 0.85) +
  labs(title = "Site representation", x = "Residue", y = "Structural divergence") +
  theme_readme

p_mode <- crop_modes(pred$mode) |>
  ggplot(aes(mode)) +
  geom_ribbon(aes(ymin = nlrmsd_msa - 1.96 * nlrmsd_msa_se,
                  ymax = nlrmsd_msa + 1.96 * nlrmsd_msa_se),
              fill = "#C41E3A", alpha = 0.18) +
  geom_line(aes(y = nlrmsd_msa), colour = "#C41E3A", linewidth = 0.6) +
  labs(title = "Mode representation", x = "Normal mode", y = NULL) +
  theme_readme

p_site | p_mode
```

`predict_decomposition()` returns more columns in the same two representations:
the profiles
of the four nested models, in `nlrmsd_mm`, `nlrmsd_ms`, `nlrmsd_ma` and
`nlrmsd_msa`; the three contributions, in `nphi_mut`, `nphi_stab` and `nphi_act`;
and a standard error for each.

```{r decomposition-call}
dec <- predict_decomposition(fit, spm, metric = "nlrmsd")

dec$site
dec$mode
```

MM, MS and MSA form a progression: MM has mutation alone, MS adds selection on
stability, and MSA adds selection on activity. Each is the previous one with one
more constraint switched on. MA, activity without stability, lies outside this
progression and is left out of the figure.

```{r nested, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3.2}
nested_long <- function(d, xvar) {
  d |>
    select({{ xvar }}, nlrmsd_mm, nlrmsd_ms, nlrmsd_msa) |>
    pivot_longer(-{{ xvar }}, names_to = "model", values_to = "nlrmsd") |>
    mutate(model = factor(model,
                          levels = c("nlrmsd_mm", "nlrmsd_ms", "nlrmsd_msa"),
                          labels = c("MM", "MS", "MSA")))
}

# Each contribution takes the colour of the model that introduces it.
nested_cols <- c(MM = "#FF8C00", MS = "#0000CD", MSA = "#C41E3A")

n_site <- nested_long(dec$site, pdb_site) |>
  ggplot(aes(pdb_site, nlrmsd, colour = model)) +
  active_lines +
  geom_line(linewidth = 0.5) +
  scale_colour_manual(values = nested_cols) +
  labs(title = "Site representation", x = "Residue", y = "Structural divergence") +
  theme_readme

n_mode <- nested_long(crop_modes(dec$mode), mode) |>
  ggplot(aes(mode, nlrmsd, colour = model)) +
  geom_line(linewidth = 0.5) +
  scale_colour_manual(values = nested_cols) +
  labs(title = "Mode representation", x = "Normal mode", y = NULL) +
  theme_readme

n_site | n_mode
```

The MSA profile splits into three contributions, one per constraint, each the
increment its step adds: mutation is MM, stability is MS minus MM, and activity is
MSA minus MS.

```{r decomposition, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3.2}
phi_long <- function(d, xvar) {
  d |>
    select({{ xvar }}, nphi_mut, nphi_stab, nphi_act) |>
    pivot_longer(-{{ xvar }}, names_to = "component", values_to = "phi") |>
    mutate(component = factor(component,
                              levels = c("nphi_mut", "nphi_stab", "nphi_act"),
                              labels = c("Mutation", "Stability", "Activity")))
}

phi_cols <- c(Mutation = "#FF8C00", Stability = "#0000CD", Activity = "#C41E3A")

d_site <- phi_long(dec$site, pdb_site) |>
  ggplot(aes(pdb_site, phi, colour = component)) +
  active_lines +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_line(linewidth = 0.5) +
  scale_colour_manual(values = phi_cols) +
  labs(title = "Site representation", x = "Residue", y = "Contribution") +
  theme_readme

d_mode <- phi_long(crop_modes(dec$mode), mode) |>
  ggplot(aes(mode, phi, colour = component)) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_line(linewidth = 0.5) +
  scale_colour_manual(values = phi_cols) +
  labs(title = "Mode representation", x = "Normal mode", y = NULL) +
  theme_readme

d_site | d_mode
```

Dashed lines mark active-site residues, and bands are 95% intervals. The mode
panels show the first `r n_modes_shown` of `r nrow(pred$mode)` modes. `nlrmsd` is
centred over all `r nrow(pred$mode)`, and these are the modes carrying the largest
divergence, so the cropped panels do not appear centred on zero.

## Reference

Echave J, Carpentier M (2026). Why structural divergence varies among residues in
enzyme evolution: contributions of mutation, stability, and activity constraints.
*Molecular Biology and Evolution* 43(7), msag162.
<https://doi.org/10.1093/molbev/msag162>

## License

MIT © Julian Echave
