---
title: "Introduction to msamodel: structure-divergence profiles and fitting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to msamodel: structure-divergence profiles and fitting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 96
)

# ggplot2 is a Suggested dependency. If it is not installed, plot chunks are
# skipped gracefully and the prose still builds.
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(eval = TRUE)

# All heavy computation (the SPM scan, the MCMC fit, the sweeps) is precomputed
# by vignettes/precompute.R and cached. Here we just load the cache and draw
# plots. The pipeline code is shown below with `eval = FALSE` so you can see how
# each cached object is produced.
cache_path <- "vignette_cache.rds"
have_cache <- file.exists(cache_path)
if (have_cache) vc <- readRDS(cache_path)
```

```{r libs, include = FALSE}
suppressMessages({
  library(msamodel)
  library(dplyr)
  library(tidyr)
})
if (has_ggplot) {
  suppressMessages(library(ggplot2))
  theme_set(theme_minimal(base_size = 12))
}

# Vignette-local goodness-of-fit helper. v0.1 does not export r2/rmse (the
# model-fit assessment layer is deferred to a later version), so we define R^2
# inline for the fitted-vs-observed plots.
r2 <- function(obs, pred) {
  1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
}
```

## What this package does

`msamodel` implements the Mutation-Stability-Activity (MSA) model of neutral
structural evolution. Given a protein structure and its active-site residues, it
predicts the **site-specific structural divergence** expected under neutral
evolution, from a single-point-mutation scan (SPM) of the structure.

The per-site quantity is the log root-mean-square structural displacement,
`lrmsd = log(sqrt(dr2_i))`, where `dr2_i` is a selection-weighted average over
the mutation scan. Two selection parameters control the weighting:

- `a1` — strength of selection on **stability**,
- `a2` — strength of selection on **activity** (proximity to the active site).

This vignette walks through everything v0.1 can do with the worked example
`1znb_A` (zinc beta-lactamase), whose data ship with the package as the `znb_*`
datasets:

```{r data}
data("znb_wt")       # elastic network model (wild type)
data("znb_spm")      # single-point-mutation scan
data("znb_profile")  # observed divergence profile (the fit target)
```

```{r show-meta, echo = FALSE, eval = have_cache}
cat("Example protein:", vc$meta$pdb_chain,
    "\nActive-site residues (PDB numbering):",
    paste(vc$meta$pdb_site_active, collapse = ", "),
    "\nSites:", vc$meta$n_sites, " | Observed sites:", vc$meta$n_obs, "\n")
```

## 1. Setup: from a PDB structure to an SPM

The shipped `znb_wt` and `znb_spm` are the cached result of the setup pipeline
below. Starting from a structure and the active-site residues, you build an
elastic network model and run a single-point-mutation scan. `setup_enm()` takes
a bio3d structure object directly, so you read the file yourself — with
`bio3d::read.pdb()` for legacy `.pdb` or `bio3d::read.cif()` for mmCIF — and the
active site is just a vector of PDB residue numbers:

```{r setup-pipeline, eval = FALSE}
# Read the structure (any source that yields a bio3d pdb object works)
pdb     <- bio3d::read.pdb(system.file("extdata", "1znb_A.pdb", package = "msamodel"))

# Active-site residues (PDB numbering)
pdb_site_active <- c(99, 101, 103, 162, 181, 184, 193, 223)

# Build the elastic network model (wild type)
wt      <- setup_enm(pdb, node = "ca", model = "ming_wall", d_max = 10.5, frustrated = FALSE)

# Single-point-mutation scan: 10 mutations per site
spm     <- generate_spm_data(wt, n_mutations = 10, model = "lfenm",
                             sigma = 0.3, min_sd = 2,
                             pdb_site_active = pdb_site_active, seed = 1024)
# `spm` is identical to the embedded `znb_spm`.
```

From the SPM we precompute the energy/displacement matrices once; every profile
below reuses them:

```{r preprocess, eval = FALSE}
pp <- preprocess_spm(znb_spm)   # list: energy_data, dr2mat, site_map
```

## 2. Divergence profiles at given (a1, a2)

For any choice of `(a1, a2)`, `calculate_dr2i_msa()` returns the per-site
weighted mean squared displacement; `lrmsd = log(sqrt(dr2_i))` is the divergence
profile. We attach two structural descriptors for context — the log RMSF
(flexibility, `log(sqrt(get_msf_site(wt)))`) and the distance to the active site
(`get_dactive`), both via `add_site_properties()`:

```{r profile-code, eval = FALSE}
a1 <- 2; a2 <- 500
profile <- calculate_dr2i_msa(pp, a1, a2) |>
  dplyr::mutate(lrmsd = log(sqrt(dr2_i)))

site_props <- add_site_properties(profile["i"], znb_wt, pdb_site_active) |>
  dplyr::mutate(lrmsf = log(sqrt(msf)))
```

The divergence profile along the chain (active sites marked in red):

```{r profile-residue, echo = FALSE, eval = have_cache && has_ggplot}
asites <- vc$meta$pdb_site_active
ggplot(vc$profile_demo, aes(pdb_site, lrmsd)) +
  geom_vline(xintercept = asites, colour = "firebrick", alpha = 0.4, linewidth = 0.3) +
  geom_line(colour = "grey40") +
  geom_point(size = 0.8, colour = "grey20") +
  labs(
    title = sprintf("Divergence profile (a1 = %g, a2 = %g)",
                    vc$meta$a1_demo, vc$meta$a2_demo),
    subtitle = "Red lines mark active-site residues",
    x = "Residue (PDB numbering)", y = "lrmsd"
  )
```

Divergence is lower (more conserved) at rigid sites and near the active site.
Plotting `lrmsd` against flexibility and against distance to the active site,
with a smooth, makes those trends explicit:

```{r profile-lrmsf, echo = FALSE, eval = have_cache && has_ggplot}
ggplot(vc$profile_demo, aes(lrmsf, lrmsd)) +
  geom_point(alpha = 0.5, colour = "grey30") +
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE, colour = "steelblue") +
  labs(title = "Divergence vs. flexibility",
       x = "lrmsf  (log RMSF)", y = "lrmsd")
```

```{r profile-dactive, echo = FALSE, eval = have_cache && has_ggplot}
ggplot(vc$profile_demo, aes(dactive, lrmsd)) +
  geom_point(alpha = 0.5, colour = "grey30") +
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE, colour = "darkorange") +
  labs(title = "Divergence vs. distance to active site",
       x = "Distance to active site (Å)", y = "lrmsd")
```

## 3. How the profile depends on a1 and a2

Stability selection (`a1`, with `a2 = 0`) and activity selection (`a2`, with
`a1 = 0`) reshape the profile in different ways. Here we vary one parameter at a
time and overlay the resulting whole (un-centered) profiles:

```{r sweep-code, eval = FALSE}
a1_values <- c(0, 1, 2, 4, 8)
sweep_a1 <- purrr::map_dfr(a1_values, function(a1)
  calculate_dr2i_msa(pp, a1, 0) |>
    dplyr::mutate(lrmsd = log(sqrt(dr2_i)), a1 = a1))
```

```{r sweep-a1, echo = FALSE, eval = have_cache && has_ggplot}
ggplot(vc$sweep_a1, aes(pdb_site, lrmsd, colour = factor(a1), group = a1)) +
  geom_vline(xintercept = vc$meta$pdb_site_active, colour = "firebrick",
             alpha = 0.3, linewidth = 0.3) +
  geom_line(alpha = 0.85) +
  scale_colour_viridis_d(end = 0.9) +
  labs(title = "Stability sweep (a2 = 0)", subtitle = "Red lines mark active-site residues",
       x = "Residue (PDB numbering)", y = "lrmsd", colour = "a1")
```

```{r sweep-a2, echo = FALSE, eval = have_cache && has_ggplot}
ggplot(vc$sweep_a2, aes(pdb_site, lrmsd, colour = factor(a2), group = a2)) +
  geom_vline(xintercept = vc$meta$pdb_site_active, colour = "firebrick",
             alpha = 0.3, linewidth = 0.3) +
  geom_line(alpha = 0.85) +
  scale_colour_viridis_d(option = "C", end = 0.9) +
  labs(title = "Activity sweep (a1 = 0)", subtitle = "Red lines mark active-site residues",
       x = "Residue (PDB numbering)", y = "lrmsd", colour = "a2")
```

Increasing `a1` suppresses divergence broadly (stronger stability selection
conserves the whole structure); increasing `a2` suppresses divergence
selectively around the active site.

## 4. Fitting a1 and a2 to observed divergence

### The four model variants

Most of what follows refers to four nested model variants, distinguished by
which selection pressures are switched on. Each is just `calculate_dr2i_msa()`
evaluated at a particular `(a1, a2)`:

| Variant | a1 (stability) | a2 (activity) | meaning |
|:--|:--|:--|:--|
| **MM** | 0 | 0 | mutation only — no selection |
| **MS** | fitted | 0 | mutation + stability selection |
| **MA** | 0 | fitted | mutation + activity selection |
| **MSA** | fitted | fitted | full model — both selections |

Their predicted profiles are `lrmsd_mm`, `lrmsd_ms`, `lrmsd_ma`, `lrmsd_msa`.
MSA is the full model fitted below; the other three isolate the effect of each
pressure and feed the decomposition in section 5.

### Running the fit

Given an observed divergence profile (`observed_data` with columns `pdb_site`
and `lrmsd_obs` — here `znb_profile`), `run_msa_bayesian_analysis()` estimates
`a1, a2` by MCMC and returns posterior summaries, per-site predictions for the
four model variants above, and the site-level decomposition of section 5:

```{r fit-code, eval = FALSE}
fit <- run_msa_bayesian_analysis(
  spm           = znb_spm,
  observed_data = znb_profile,
  n_mcmc_iter   = 4000,   # for demonstration; use more for production
  n_burnin      = 1000
)
fit$parameter_summary
```

```{r fit-params, echo = FALSE, eval = have_cache}
knitr::kable(vc$parameter_summary, digits = 3,
             caption = "Posterior summary of the selection parameters")
```

### Fitted vs. observed

The likelihood compares **mean-centered** profiles (`nlrmsd = lrmsd - mean(lrmsd)`),
so the comparison plots use the same centering. We annotate each with the
coefficient of determination R² between centered observed and centered fitted
values:

```{r fit-prep, include = FALSE, eval = have_cache}
cmp <- vc$comparison
r2_fit <- r2(cmp$nlrmsd_obs, cmp$nlrmsd_fit)
cmp_long <- cmp |>
  tidyr::pivot_longer(c(nlrmsd_obs, nlrmsd_fit),
                      names_to = "source", values_to = "nlrmsd") |>
  dplyr::mutate(source = dplyr::recode(source,
                                       nlrmsd_obs = "observed",
                                       nlrmsd_fit = "fitted (MSA)"))
```

Profile along the chain — observed vs fitted:

```{r fit-profile, echo = FALSE, eval = have_cache && has_ggplot}
ggplot(cmp_long, aes(pdb_site, nlrmsd, colour = source)) +
  geom_line(alpha = 0.8) +
  scale_colour_manual(values = c("observed" = "grey25", "fitted (MSA)" = "firebrick")) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.15, vjust = 1.8,
           label = sprintf("R^2 == %.3f", r2_fit), parse = TRUE) +
  labs(title = "Observed vs. fitted divergence profile",
       x = "Residue (PDB numbering)", y = "nlrmsd (mean-centered)", colour = NULL) +
  theme(legend.position = "top")
```

Fitted vs observed as a scatter (the gof view):

```{r fit-scatter, echo = FALSE, eval = have_cache && has_ggplot}
ggplot(cmp, aes(nlrmsd_obs, nlrmsd_fit)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linetype = 2) +
  geom_point(alpha = 0.5, colour = "steelblue") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.15, vjust = 1.4,
           label = sprintf("R^2 == %.3f", r2_fit), parse = TRUE) +
  labs(title = "Fitted vs. observed (mean-centered)",
       x = "observed nlrmsd", y = "fitted nlrmsd")
```

And the fitted/observed divergence against flexibility and distance to the
active site:

```{r fit-vs-lrmsf, echo = FALSE, eval = have_cache && has_ggplot}
ggplot(cmp_long, aes(lrmsf, nlrmsd, colour = source)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  scale_colour_manual(values = c("observed" = "grey25", "fitted (MSA)" = "firebrick")) +
  labs(title = "Observed vs. fitted, by flexibility",
       x = "lrmsf (log RMSF)", y = "nlrmsd (mean-centered)", colour = NULL) +
  theme(legend.position = "top")
```

```{r fit-vs-dactive, echo = FALSE, eval = have_cache && has_ggplot}
ggplot(cmp_long, aes(dactive, nlrmsd, colour = source)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  scale_colour_manual(values = c("observed" = "grey25", "fitted (MSA)" = "firebrick")) +
  labs(title = "Observed vs. fitted, by distance to active site",
       x = "Distance to active site (Å)", y = "nlrmsd (mean-centered)", colour = NULL) +
  theme(legend.position = "top")
```

## 5. Site-level decomposition

The fit also splits each site's full predicted divergence (`lrmsd_msa`) into
three additive contributions — a mutation term, a stability-selection term, and
an activity-selection term — built from the four model variants of section 4.
Each selection term is the *average* of its marginal effect measured two ways:
added on its own to the mutation-only model, and added last to the other
single-selection model. Writing $\phi$ for the contributions,

$$
\phi_{\text{mut}}  = \text{lrmsd}_{\text{mm}}
$$
$$
\phi_{\text{stab}} = \tfrac{1}{2}\bigl[(\text{lrmsd}_{\text{ms}} - \text{lrmsd}_{\text{mm}}) + (\text{lrmsd}_{\text{msa}} - \text{lrmsd}_{\text{ma}})\bigr]
$$
$$
\phi_{\text{act}}  = \tfrac{1}{2}\bigl[(\text{lrmsd}_{\text{ma}} - \text{lrmsd}_{\text{mm}}) + (\text{lrmsd}_{\text{msa}} - \text{lrmsd}_{\text{ms}})\bigr]
$$

By construction these sum to the full prediction,
$\phi_{\text{mut}} + \phi_{\text{stab}} + \phi_{\text{act}} = \text{lrmsd}_{\text{msa}}$.
The mutation term is the divergence under no selection (it still varies among
sites — mutations perturb sites unequally); a negative selection term means that
pressure *reduces* divergence at that site relative to the mutation-only model.
`run_msa_bayesian_analysis()` returns these as `phi_mut`, `phi_stab`,
`phi_act` in `decomposition_summary` (one posterior summary per site and term).

```{r decomp, echo = FALSE, eval = have_cache && has_ggplot}
labs_comp <- c(phi_mut = "mutation",
               phi_stab = "stability selection",
               phi_act = "activity selection")
dec <- vc$decomposition_summary |>
  dplyr::mutate(component = factor(labs_comp[component], levels = labs_comp))
ggplot(dec, aes(pdb_site, mean)) +
  geom_vline(xintercept = vc$meta$pdb_site_active, colour = "firebrick",
             alpha = 0.3, linewidth = 0.3) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_line(colour = "grey30") +
  facet_wrap(~ component, ncol = 1, scales = "free_y") +
  labs(title = "Decomposition of predicted divergence",
       subtitle = "Red lines mark active-site residues",
       x = "Residue (PDB numbering)", y = "contribution to lrmsd")
```

## 6. Systematic 2-D scans of (a1, a2)

The one-at-a-time sweeps of section 3 generalise to a full 2-D scan with no extra
machinery: preprocess the SPM **once** (the reshape is independent of `a1, a2`),
then map `calculate_dr2i_msa()` over a grid of parameter pairs. Reusing the same
`pp` for every point is the whole point — re-preprocessing per point would be
wasted work.

```{r grid-code, eval = FALSE}
pp   <- preprocess_spm(znb_spm)            # reshape once, reuse for every (a1, a2)
grid <- expand.grid(a1 = c(0, 2, 4, 8),
                    a2 = c(0, 31, 127, 511))
scan <- purrr::map_dfr(seq_len(nrow(grid)), function(k) {
  calculate_dr2i_msa(pp, grid$a1[k], grid$a2[k]) |>
    dplyr::mutate(a1 = grid$a1[k], a2 = grid$a2[k])
})
```

## Session info

```{r session, echo = FALSE}
sessionInfo()
```
