---
title: "Fitting the model to data, with confidence bands"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting the model to data, with confidence bands}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 3.6,
  dpi = 96
)

# Every chunk below runs for real when this file is knit.
# The fit is ML (deterministic); the seed is harmless but kept for a stable render.
set.seed(1024)
```

```{r libs, include = FALSE}
suppressMessages({
  library(msamodel)
  library(penm)
  library(dplyr)
  library(tidyr)
})
suppressMessages(library(ggplot2))
suppressMessages(library(patchwork))
theme_set(theme_minimal(base_size = 11))
```

The [explore vignette](msamodel-explore.html) evaluated the model at `(a1, a2)` values you
choose. This one **estimates** them from a divergence profile you measured, and reports
every prediction with a confidence band. For a short end-to-end tour, see the
[package overview](msamodel.html).

A point estimate without an uncertainty is not worth much, so each prediction here comes
with a standard error, and the band is the value plus or minus a couple of standard
errors. That standard error combines two sources: the fit knows `(a1, a2)` only to within
its own uncertainty, and a small wobble there produces a small wobble in every predicted
value; and the mutation scan behind the prediction is finite, which adds its own sampling
error. The reported `_se` already combines both. The fit is deterministic — a numerical
optimisation, no chain and no burn-in — so the same data always gives the same answer.

**You fit on one axis and read the prediction on both.** The observed data is per
residue, so `fit_lrmsd_msa_site()` is what estimates `(a1, a2)`. But the resulting fit
describes the model, not the axis: `predict_profiles()` and `predict_decomposition()`
each return a `$site` and a `$mode` tibble from that one fit. Every prediction below is
shown side by side — the residues you fitted, and the collective motions implied by the
same parameters.

## One idea to grasp first: `lrmsd` vs. `nlrmsd`

Each prediction verb takes a `metric` argument selecting one of two forms, and knowing
which to reach for is the one thing worth understanding before you start.

- **`lrmsd`** (`metric = "lrmsd"`) is the profile the model predicts directly: the log of
  each residue's (or mode's) structural divergence. The raw predicted curve.
- **`nlrmsd`** (`metric = "nlrmsd"`) is that same curve with its average subtracted off:
  `nlrmsd = lrmsd - mean(lrmsd)`. It describes the *shape* of the profile — which
  residues are above or below average — with the overall level removed. The leading `n`
  marks this centred form.

Why keep both? Because they answer different questions, and each carries its **own**
confidence band. It is tempting to assume `nlrmsd`'s band is just `lrmsd`'s band shifted
down by the mean — but it is genuinely narrower. The reason: the mean you subtract is
itself computed from the model, so it moves with `(a1, a2)` too. When `(a1, a2)` wobble,
the whole curve and its mean wobble *together*, and part of that wobble cancels in
`lrmsd - mean(lrmsd)`. The shape is thus more tightly pinned down than the raw curve.
These are two real predictions with two real uncertainties, not one prediction drawn
twice.

**Which to use:** compare against measured data with **`nlrmsd`**. A real divergence
profile can only be compared up to an overall level — the data does not pin down the
absolute scale — so the meaningful comparison is of shape, exactly what `nlrmsd` holds.
Reach for **`lrmsd`** when you want the model's raw predicted curve on its own. The
decomposition exists only in the centred form, because there is no meaningful way to
split up an overall level the data cannot see.

## From a standard error to a band

Every prediction verb reports each value with a `_se` sibling column. To draw a
confidence band, turn it into lower/upper bounds: a 95% band is the value plus or minus
`z_975` standard errors, with `z_975 = qnorm(0.975) ≈ 1.96` (for a 99% band use
`qnorm(0.995) ≈ 2.58`). We define the multiplier once and reuse it throughout:

```{r zconst}
z_975 <- qnorm(0.975)   # 1.96; the multiplier for a 95% band
```

## 1. Setup

The starting point is the same as in the [explore vignette](msamodel-explore.html): a
structure file, a vector of active-site residue numbers, an elastic network model, and a
mutation scan. Swap in your own files at the two marked lines.

```{r attach}
library(msamodel)
library(penm)       # penm::set_enm(), the elastic network model msamodel builds on
library(dplyr)      # data-wrangling verbs (mutate, inner_join, ...) used below
library(tidyr)      # pivot_longer, to reshape profiles for plotting
library(readr)      # read_csv, for the example data files
library(ggplot2)    # the figures
library(patchwork)  # to place the site and mode panels side by side
```

```{r inputs}
pdb_file    <- system.file("extdata", "1d6o_A.pdb", package = "msamodel")          # <- your PDB
active_file <- system.file("extdata", "1d6o_A_active_site.csv", package = "msamodel") # <- your active sites

pdb_site_active <- read_csv(active_file, show_col_types = FALSE) |>
  filter(pdb == "1d6o", chain == "A") |>
  pull(pdb_site)
```

```{r build}
wt_pdb <- bio3d::read.pdb(pdb_file)
wt     <- penm::set_enm(wt_pdb, node = "ca", model = "ming_wall", d_max = 10.5, frustrated = FALSE)
spm    <- generate_spm(wt, n_mutations = 10, model = "lfenm", sigma = 0.3,
                            min_sd = 2, pdb_site_active = pdb_site_active, ensemble = 1L)
```

## 2. Your observed data, and the fit

To fit, you supply an **observed** divergence profile — **your** data, not something
`msamodel` computes. You obtain it upstream by superimposing a set of homologous
structures onto your protein over a sequence alignment and measuring, at each residue,
how much its position varies across the family; the observed value is the log of that
per-residue variation. `msamodel` does not do this step — you bring the result as **two
vectors**: the PDB residue numbers, and the observed value at each. They can live in a
table of any shape, with columns named whatever you already call them.

```{r observed}
observed_file <- system.file("extdata", "1d6o_A_lrmsd_obs_site.csv", package = "msamodel")
observed      <- read_csv(observed_file, show_col_types = FALSE)   # <- your observed data
head(observed)
```

An observed profile need not cover every residue the model spans — an alignment often
measures only some of them — and the fitter simply uses the overlap. This example happens
to cover all 107, but nothing requires that of your own data.

`fit_lrmsd_msa_site()` estimates `(a1, a2)` by maximum likelihood, returning point
estimates plus an asymptotic covariance from the Hessian at the optimum:

```{r fit}
fit <- fit_lrmsd_msa_site(spm, observed$pdb_site, observed$lrmsd_obs)
c(a1 = fit$a1, a2 = fit$a2)
```

```{r fit-params, echo = FALSE}
knitr::kable(
  tibble::tibble(
    parameter = c("a1", "a2"),
    estimate  = c(fit$a1, fit$a2),
    se        = c(fit$se_a1, fit$se_a2)
  ),
  digits = 3,
  caption = "ML estimate of the selection parameters (se from the Hessian)"
)
```

### How well does it fit?

Goodness of fit comes back attached to the fit object, as `fit$gof`, so scoring is not a
separate call:

```{r gof}
fit$gof
```

We score with `D²`, the deviance explained relative to a flat (mean-only) null:
`D² = 1 - deviance/null_deviance`, computed over exactly the residues the fitter scored.
Note it is `D²`, not `R²`: the model predicts divergence on a fixed scale, and `D²`
measures agreement with the prediction *as it stands*, where `R²` would credit a
prediction that is right only up to an arbitrary rescaling. `D²` is at most 1 but has no
lower bound — a negative value means the fit is worse than a constant profile.

## 3. The predicted profile, with a band

`predict_profiles()` evaluates the model at the fitted `(a1, a2)` and returns both
readings, each with its standard error:

```{r predict-profiles}
predicted <- predict_profiles(fit, spm, metric = "nlrmsd")
head(predicted$site)
head(predicted$mode)
```

We turn the standard errors into 95% bands, and centre the observed profile the same way
the prediction is centred, so the two are on one scale:

```{r bands}
add_band <- function(d) {
  d |> mutate(lower = nlrmsd_msa - z_975 * nlrmsd_msa_se,
              upper = nlrmsd_msa + z_975 * nlrmsd_msa_se)
}

profile_site <- predicted$site |>
  add_band() |>
  inner_join(observed, by = "pdb_site") |>
  mutate(nlrmsd_obs = lrmsd_obs - mean(lrmsd_obs))

profile_mode <- add_band(predicted$mode)
```

On the **site** axis the prediction can be compared against your data, because that is
where the data lives. On the **mode** axis there is nothing to compare against: the panel
is the same fit read in the other basis, a prediction about which collective motions
carry the divergence. That asymmetry holds for the rest of the vignette.

```{r profile-panels}
pp_site <- ggplot(profile_site, aes(pdb_site)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "firebrick", alpha = 0.2) +
  geom_line(aes(y = nlrmsd_msa, colour = "fitted (MSA)")) +
  geom_line(aes(y = nlrmsd_obs, colour = "observed"), alpha = 0.8) +
  scale_colour_manual(values = c("observed" = "grey25", "fitted (MSA)" = "firebrick")) +
  labs(subtitle = "by residue: fitted vs. observed",
       x = "Residue (PDB numbering)", y = "nlrmsd (shape)", colour = NULL) +
  theme(legend.position = "bottom")

pp_mode <- ggplot(profile_mode, aes(mode)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey60", alpha = 0.35) +
  geom_line(aes(y = nlrmsd_msa), colour = "grey30") +
  labs(subtitle = "by normal mode: predicted", x = "Normal-mode index", y = NULL)

pp_site + pp_mode +
  plot_annotation(title = "Predicted divergence profile at the ML estimate",
                  subtitle = "Shaded: 95% band from the reported standard errors")
```

Fitted against observed as a scatter — points on the dashed line are a perfect match:

```{r fit-scatter, fig.width = 4.6, fig.height = 4.2}
ggplot(profile_site, aes(nlrmsd_obs, nlrmsd_msa)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linetype = 2) +
  geom_point(alpha = 0.5, colour = "steelblue") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.15, vjust = 1.4,
           label = sprintf("D^2 == %.3f", fit$gof$D2), parse = TRUE) +
  labs(title = "Fitted vs. observed", x = "observed nlrmsd", y = "fitted nlrmsd")
```

## 4. The nested models, with bands

The same four nested variants the [explore vignette](msamodel-explore.html) introduced — MM
(no selection), MS (stability only), MA (activity only), MSA (both) — come back from
`predict_decomposition()`, now each with its own standard error:

```{r predict-decomposition}
decomposition <- predict_decomposition(fit, spm, metric = "nlrmsd")
names(decomposition$site)
```

Every value column has a `_se` sibling, so any of them can be banded the same way. Here
we band all four nested profiles and show them together:

```{r nested-panels, fig.height = 4.6}
model_colours <- c(MM = "#FF8C00", MS = "#0000CD", MA = "#1B9E77", MSA = "#C41E3A")

# `key` is the axis column name ("pdb_site" or "mode"): the value columns and the
# _se columns are pivoted separately, then joined back on that key plus the model.
to_nested_long <- function(d, key) {
  values <- d |>
    select(all_of(key), nlrmsd_mm, nlrmsd_ms, nlrmsd_ma, nlrmsd_msa) |>
    pivot_longer(-all_of(key), names_to = "model", values_to = "nlrmsd")
  errors <- d |>
    select(all_of(key), nlrmsd_mm_se, nlrmsd_ms_se, nlrmsd_ma_se, nlrmsd_msa_se) |>
    pivot_longer(-all_of(key), names_to = "model", values_to = "se") |>
    mutate(model = sub("_se$", "", model))
  values |>
    left_join(errors, by = c(key, "model")) |>
    mutate(model = factor(toupper(sub("nlrmsd_", "", model)),
                          levels = c("MM", "MS", "MA", "MSA")),
           lower = nlrmsd - z_975 * se,
           upper = nlrmsd + z_975 * se)
}

nested_site <- to_nested_long(decomposition$site, "pdb_site")
nested_mode <- to_nested_long(decomposition$mode, "mode")

nn_site <- ggplot(nested_site, aes(pdb_site, nlrmsd, colour = model, fill = model)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, colour = NA) +
  geom_line(alpha = 0.9) +
  scale_colour_manual(values = model_colours) +
  scale_fill_manual(values = model_colours) +
  labs(subtitle = "by residue", x = "Residue (PDB numbering)", y = "nlrmsd") +
  theme(legend.position = "bottom")

nn_mode <- ggplot(nested_mode, aes(mode, nlrmsd, colour = model, fill = model)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, colour = NA) +
  geom_line(alpha = 0.9) +
  scale_colour_manual(values = model_colours, guide = "none") +
  scale_fill_manual(values = model_colours, guide = "none") +
  labs(subtitle = "by normal mode", x = "Normal-mode index", y = NULL)

nn_site + nn_mode +
  plot_annotation(title = "Nested-model profiles at the ML estimate, with 95% bands")
```

Each model adds one more selection pressure. Where the bands of two variants overlap, the
data does not distinguish them at that position.

## 5. The decomposition, with bands

The same call carries the mutation / stability / activity split. In the centred form the
columns are `nphi_mut`, `nphi_stab`, `nphi_act`, each with its `_se`:

```{r decomposition-panels, fig.height = 5.6}
component_colours <- c(nphi_mut = "#FF8C00", nphi_stab = "#0000CD", nphi_act = "#C41E3A")

to_phi_long <- function(d, key) {
  values <- d |>
    select(all_of(key), nphi_mut, nphi_stab, nphi_act) |>
    pivot_longer(-all_of(key), names_to = "component", values_to = "phi")
  errors <- d |>
    select(all_of(key), nphi_mut_se, nphi_stab_se, nphi_act_se) |>
    pivot_longer(-all_of(key), names_to = "component", values_to = "se") |>
    mutate(component = sub("_se$", "", component))
  values |>
    left_join(errors, by = c(key, "component")) |>
    mutate(component = factor(component, levels = names(component_colours)),
           lower = phi - z_975 * se,
           upper = phi + z_975 * se)
}

dd_site <- ggplot(to_phi_long(decomposition$site, "pdb_site"),
                  aes(pdb_site, phi, colour = component, fill = component)) +
  geom_vline(xintercept = pdb_site_active, colour = "firebrick",
             alpha = 0.3, linewidth = 0.3) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  geom_line() +
  scale_colour_manual(values = component_colours, guide = "none") +
  scale_fill_manual(values = component_colours, guide = "none") +
  facet_wrap(~ component, ncol = 1) +
  labs(subtitle = "by residue", x = "Residue (PDB numbering)", y = "contribution to nlrmsd")

dd_mode <- ggplot(to_phi_long(decomposition$mode, "mode"),
                  aes(mode, phi, colour = component, fill = component)) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  geom_line() +
  scale_colour_manual(values = component_colours, guide = "none") +
  scale_fill_manual(values = component_colours, guide = "none") +
  facet_wrap(~ component, ncol = 1, scales = "free_y") +
  labs(subtitle = "by normal mode", x = "Normal-mode index", y = NULL)

dd_site + dd_mode +
  plot_annotation(title = "Decomposition at the ML estimate, with 95% bands",
                  subtitle = "Left: red lines mark active-site residues")
```

Each curve is one pressure's contribution at that residue or mode. Where a curve dips
**below** zero, that pressure reduces divergence there relative to the mutation-only
model; above zero, it adds divergence. A band that straddles zero means the data does not
resolve that contribution's sign at that position — the reason for carrying the standard
errors through the decomposition rather than reporting the split alone.

## Where to next

- **[Explore](msamodel-explore.html)** — how the profiles and their contributions respond to
  `a1` and `a2` when you set the parameters yourself.
- **[Package overview](msamodel.html)** — the short end-to-end tour.

## Session info

```{r session, echo = FALSE}
sessionInfo()
```
