---
title: "SPM sampling-error band: delta method vs bootstrap"
author: "msamodel dev validation"
output:
  pdf_document:
    toc: true
    number_sections: true
    df_print: kable
geometry: margin=1in
header-includes:
  - \usepackage{amsmath}
  - \usepackage{float}
---

```{r setup, include=FALSE}
# fig.width just under the text block (6.5in at 1in margins) so the rightmost facet's
# outermost x tick label never touches the page edge; figures pinned with fig.pos.
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 6.2, fig.height = 4.4, out.width = "100%",
                      fig.align = "center", fig.pos = "H", dpi = 130)
suppressMessages({
  devtools::load_all(quiet = TRUE)
  library(ggplot2); library(dplyr); library(tidyr)
})
set.seed(20260724)
```

# What this document validates, and why it exists

The `predict_*_ml` band functions report a **finite-mutation sampling error** (the "SPM
arm") in addition to the `(a1, a2)` parameter-uncertainty arm. The SPM arm is computed by
a delta-method formula. This document is the durable, reproducible check that that formula
is correct: it recomputes every SPM band from the **shipped package internals** and
compares it, per residue, to an **independent bootstrap** — a genuinely different
computation (resampling), so it can disagree when the delta formula is wrong.

Knit with

```r
rmarkdown::render("dev/reports/spm_band_validation.Rmd")
```

# The quantities

Let $M$ be the SPM divergence matrix `dr2_ijm` (rows $k$ = mutants, columns $i$ =
response sites). For selection strengths $(a_1,a_2)$ the model assigns each mutant an
unnormalised fixation probability $p_k$ (`pfix_msa`), normalised to weights
$w_k = p_k/\sum_k p_k$. The model's predicted per-site divergence is the self-normalised
weighted mean, and the fit works in its log and mean-centred forms:

\begin{equation}
\mathrm{msd}_i \;=\; \sum_k w_k\, M_{ki},
\qquad
\mathrm{lrmsd}_i \;=\; \tfrac12 \log \mathrm{msd}_i,
\qquad
\mathrm{nlrmsd}_i \;=\; \mathrm{lrmsd}_i - \frac{1}{|S|}\sum_{s\in S}\mathrm{lrmsd}_s ,
\end{equation}

the last being the mean-centred quantity (over the response support $S$) that the fit
works on. The four nested variants MM/MS/MA/MSA are the same formulas at
$(0,0)/(a_1,0)/(0,a_2)/(a_1,a_2)$; the decomposition contributions are the contrasts
$\phi_{\mathrm{mut}}=\mathrm{nlrmsd}_{\mathrm{mm}}$,
$\phi_{\mathrm{stab}}=\mathrm{nlrmsd}_{\mathrm{ms}}-\mathrm{nlrmsd}_{\mathrm{mm}}$,
$\phi_{\mathrm{act}}=\mathrm{nlrmsd}_{\mathrm{msa}}-\mathrm{nlrmsd}_{\mathrm{ms}}$.

Only the finite mutant sample is random here (parameters are held fixed); the band is the
sampling standard deviation of these quantities.

# Delta method (the shipped formula)

The package computes a **per-mutant contribution matrix** $h$ (internal `spm_hmat`). With
$G_{ki} = M_{ki}/(2\,\mathrm{msd}_i)$ the log-sensitivity (the derivative of
$\tfrac12\log\mathrm{msd}$):

\begin{align}
G_{ki} &= \frac{M_{ki}}{2\,\mathrm{msd}_i}, &
&\text{(if centred: } G_{ki} \leftarrow G_{ki} - \tfrac{1}{|S|}\textstyle\sum_s G_{ks}\text{)} \\
h_{ki} &= w_k\Big( G_{ki} - \sum_{k'} w_{k'} G_{k'i} \Big).
\end{align}

The centring subtraction $G-\operatorname{rowMeans}(G)$ is $\partial\,\mathrm{nlrmsd}/\partial M$:
each mutant drives the support mean too. Then

\begin{itemize}
\item \textbf{Profile band} (single model): $\displaystyle \mathrm{SE}(\mathrm{lrmsd}_i) = \sqrt{\textstyle\sum_k h_{ki}^2}$.
\item \textbf{Contrast band} ($\phi = \mathrm{lrmsd}_A-\mathrm{lrmsd}_B$ on the \emph{same} scan):
      $\displaystyle \mathrm{SE} = \sqrt{\textstyle\sum_k (h^A_{ki}-h^B_{ki})^2}$.
\end{itemize}

The contrast's shared-sample cross term is automatic: the same mutant $k$ enters both
$h^A$ and $h^B$, so $\sum_k(h^A-h^B)^2 = \sum_k h_A^2 + \sum_k h_B^2 - 2\sum_k h_A h_B$,
the last being $-2\,\widehat{\mathrm{Cov}}$. Equivalently, for the uncentred single model
the formula reduces to the recorded

\begin{equation}
\mathrm{SE}(\mathrm{lrmsd}_i) \;=\; \frac{1}{2\,\mathrm{msd}_i}\sqrt{\sum_k w_k^2\,(M_{ki}-\mathrm{msd}_i)^2},
\end{equation}

since the $w$-weighted mean of $G_{\cdot i}$ is exactly $\tfrac12$.

`Var = SE^2` is what the predictors add to the parameter-arm variance (under
independence) before forming the band.

# Bootstrap (the independent reference)

This is the reference the delta formula is checked against — the correct bootstrap for
this problem, not a simplified one. Per replicate: draw $N$ mutant rows **uniformly with
replacement** — this is how the scan drew them; the model weights are *not* sampling
probabilities. Drawing with probability
proportional to $w$ would double-count the weights (once in the draw, once in the mean)
and inflate the profile SD by about 10% for weighted ensembles (it vanishes only for
uniform MM). On the
drawn multiset, recompute each model's Hájek weighted mean

\begin{equation}
\widetilde{\mathrm{msd}}_i^{(v)} \;=\; \frac{\sum_{k\in\text{draw}} w^{(v)}_k M_{ki}}{\sum_{k\in\text{draw}} w^{(v)}_k},
\end{equation}

take $\tfrac12\log$, centre, and difference (a **shared** draw across variants preserves
the cross-model correlation for $\phi$). The band is the SD across $B$ replicates.

```{r compute}
pp <- znb_spm; M <- pp$dr2_ijm
N  <- nrow(M); S <- ncol(M); ed <- pp$energy_data
ml <- fit_lrmsd_msa_site(pp, znb_profile)
a1 <- ml$a1; a2 <- ml$a2
B  <- 2000

Wp <- list(MM  = pfix_msa(ed$ddg_jm, ed$ddgact_jm, 0, 0),
           MS  = pfix_msa(ed$ddg_jm, ed$ddgact_jm, a1, 0),
           MA  = pfix_msa(ed$ddg_jm, ed$ddgact_jm, 0, a2),
           MSA = pfix_msa(ed$ddg_jm, ed$ddgact_jm, a1, a2))
neff <- function(p) sum(p)^2 / sum(p^2)

# DELTA via shipped internals: h-matrix (centred/uncentred) per variant
h_u <- lapply(Wp, function(w) spm_hmat(M, w, centred = FALSE))
h_c <- lapply(Wp, function(w) spm_hmat(M, w, centred = TRUE))
delta <- list()
for (v in names(Wp)) {
  delta[[paste0(v, " (uncentred)")]] <- sqrt(colSums(h_u[[v]]^2))
  delta[[paste0(v, " (centred)")]]   <- sqrt(colSums(h_c[[v]]^2))
}
delta[["phi_mut (centred)"]]  <- sqrt(colSums(h_c$MM^2))
delta[["phi_stab (centred)"]] <- sqrt(colSums((h_c$MS  - h_c$MM)^2))
delta[["phi_act (centred)"]]  <- sqrt(colSums((h_c$MSA - h_c$MS)^2))

# BOOTSTRAP: shared uniform draw
accU <- lapply(Wp, function(w) matrix(NA_real_, B, S))
accC <- lapply(Wp, function(w) matrix(NA_real_, B, S))
Pstab <- matrix(NA_real_, B, S); Pact <- matrix(NA_real_, B, S); Pmut <- matrix(NA_real_, B, S)
for (b in seq_len(B)) {
  idx <- sample.int(N, N, replace = TRUE); sub <- M[idx, , drop = FALSE]
  lc <- list()
  for (v in names(Wp)) {
    wv <- Wp[[v]][idx]; wv <- wv / sum(wv)
    lb <- 0.5 * log(colSums(sub * wv))
    accU[[v]][b, ] <- lb; accC[[v]][b, ] <- lb - mean(lb); lc[[v]] <- lb - mean(lb)
  }
  Pmut[b, ]  <- lc$MM
  Pstab[b, ] <- lc$MS  - lc$MM
  Pact[b, ]  <- lc$MSA - lc$MS
}
sdcol <- function(m) apply(m, 2, stats::sd)
boot <- list()
for (v in names(Wp)) {
  boot[[paste0(v, " (uncentred)")]] <- sdcol(accU[[v]])
  boot[[paste0(v, " (centred)")]]   <- sdcol(accC[[v]])
}
boot[["phi_mut (centred)"]]  <- sdcol(Pmut)
boot[["phi_stab (centred)"]] <- sdcol(Pstab)
boot[["phi_act (centred)"]]  <- sdcol(Pact)

quantities <- names(delta)
dat <- bind_rows(lapply(quantities, function(f)
  tibble(quantity = f, site = seq_len(S), delta_sd = delta[[f]], boot_sd = boot[[f]])))
dat$quantity <- factor(dat$quantity, levels = quantities)
dat$group  <- ifelse(grepl("^phi", dat$quantity), "phi decomposition", "profiles")
```

\clearpage

# Results — site axis

```{r summary}
summ <- dat %>% group_by(quantity) %>%
  summarise(`median delta/boot` = median(delta_sd / boot_sd),
            q10 = quantile(delta_sd/boot_sd, .1),
            q90 = quantile(delta_sd/boot_sd, .9),
            `min delta SD` = min(delta_sd), .groups = "drop")
knitr::kable(summ, digits = 3,
             caption = "Site axis: per-quantity agreement. Ratio near 1 across all 11 quantities.")
```

Effective sample sizes $N_{\text{eff}}=(\sum p)^2/\sum p^2$:
`r sprintf("MSA=%.0f, MS=%.0f, MA=%.0f, MM=%.0f (N=%d)", neff(Wp$MSA), neff(Wp$MS), neff(Wp$MA), neff(Wp$MM), N)`.
The weighted ensembles use about 28% of the mutants effectively, so the between-model
correlation is large — which is why the $\phi$ contrast **must** keep the $-2\,\mathrm{Cov}$
cross term (a variance-sum formula is off by roughly 3x here).

```{r scatter, fig.height=6.6, fig.cap="Site axis: per-site delta SD vs bootstrap SD. Points on the y=x line = agreement."}
ggplot(dat, aes(boot_sd, delta_sd, colour = group)) +
  geom_abline(slope = 1, colour = "grey60") +
  geom_point(alpha = .5, size = .7) +
  facet_wrap(~quantity, scales = "free", ncol = 3) +
  scale_colour_manual(values = c("phi decomposition" = "#c1442e", profiles = "#2e6fc1")) +
  labs(x = "bootstrap SD (uniform-resample)", y = "delta SD", colour = NULL) +
  theme_minimal(9) + theme(legend.position = "top")
```

```{r ratio, fig.height=6.6, fig.cap="Site axis: per-site ratio delta/bootstrap (1.0 = agreement)."}
ggplot(dat, aes(site, delta_sd / boot_sd, colour = group)) +
  geom_hline(yintercept = 1, colour = "grey60") +
  geom_point(alpha = .5, size = .7) +
  facet_wrap(~quantity, scales = "free_y", ncol = 3) +
  scale_colour_manual(values = c("phi decomposition" = "#c1442e", profiles = "#2e6fc1")) +
  coord_cartesian(ylim = c(0.7, 1.3)) +
  labs(x = "site", y = "delta / bootstrap", colour = NULL) +
  theme_minimal(9) + theme(legend.position = "top")
```

\clearpage

# Results — mode axis

The mode axis is the exact parallel of the site axis: the same delta and bootstrap
procedures, on the per-mode divergence matrix `dr2_njm` (from `preprocess_spm_mode`)
instead of the per-site `dr2_ijm`, and the mode fit `fit_lrmsd_msa_mode`. It is validated
separately because the two axes have separate preprocessing and response matrices.

```{r compute-mode}
ppm <- preprocess_spm_mode(znb_spm); Mn <- ppm$dr2_njm
Nn  <- nrow(Mn); Sn <- ncol(Mn); edn <- ppm$energy_data
mln <- fit_lrmsd_msa_mode(ppm, znb_profile_n)
a1n <- mln$a1; a2n <- mln$a2

Wn <- list(MM  = pfix_msa(edn$ddg_jm, edn$ddgact_jm, 0, 0),
           MS  = pfix_msa(edn$ddg_jm, edn$ddgact_jm, a1n, 0),
           MA  = pfix_msa(edn$ddg_jm, edn$ddgact_jm, 0, a2n),
           MSA = pfix_msa(edn$ddg_jm, edn$ddgact_jm, a1n, a2n))

# DELTA via shipped internals on the mode matrix
hn_u <- lapply(Wn, function(w) spm_hmat(Mn, w, centred = FALSE))
hn_c <- lapply(Wn, function(w) spm_hmat(Mn, w, centred = TRUE))
delta_n <- list()
for (v in names(Wn)) {
  delta_n[[paste0(v, " (uncentred)")]] <- sqrt(colSums(hn_u[[v]]^2))
  delta_n[[paste0(v, " (centred)")]]   <- sqrt(colSums(hn_c[[v]]^2))
}
delta_n[["phi_mut (centred)"]]  <- sqrt(colSums(hn_c$MM^2))
delta_n[["phi_stab (centred)"]] <- sqrt(colSums((hn_c$MS  - hn_c$MM)^2))
delta_n[["phi_act (centred)"]]  <- sqrt(colSums((hn_c$MSA - hn_c$MS)^2))

# BOOTSTRAP: shared uniform draw over the same mutant rows
accUn <- lapply(Wn, function(w) matrix(NA_real_, B, Sn))
accCn <- lapply(Wn, function(w) matrix(NA_real_, B, Sn))
Pstabn <- matrix(NA_real_, B, Sn); Pactn <- matrix(NA_real_, B, Sn); Pmutn <- matrix(NA_real_, B, Sn)
for (b in seq_len(B)) {
  idx <- sample.int(Nn, Nn, replace = TRUE); sub <- Mn[idx, , drop = FALSE]
  lc <- list()
  for (v in names(Wn)) {
    wv <- Wn[[v]][idx]; wv <- wv / sum(wv)
    lb <- 0.5 * log(colSums(sub * wv))
    accUn[[v]][b, ] <- lb; accCn[[v]][b, ] <- lb - mean(lb); lc[[v]] <- lb - mean(lb)
  }
  Pmutn[b, ]  <- lc$MM
  Pstabn[b, ] <- lc$MS  - lc$MM
  Pactn[b, ]  <- lc$MSA - lc$MS
}
bootn <- list()
for (v in names(Wn)) {
  bootn[[paste0(v, " (uncentred)")]] <- sdcol(accUn[[v]])
  bootn[[paste0(v, " (centred)")]]   <- sdcol(accCn[[v]])
}
bootn[["phi_mut (centred)"]]  <- sdcol(Pmutn)
bootn[["phi_stab (centred)"]] <- sdcol(Pstabn)
bootn[["phi_act (centred)"]]  <- sdcol(Pactn)

datn <- bind_rows(lapply(quantities, function(f)
  tibble(quantity = f, mode = seq_len(Sn), delta_sd = delta_n[[f]], boot_sd = bootn[[f]])))
datn$quantity <- factor(datn$quantity, levels = quantities)
datn$group  <- ifelse(grepl("^phi", datn$quantity), "phi decomposition", "profiles")
```

```{r summary-mode}
summ_n <- datn %>% group_by(quantity) %>%
  summarise(`median delta/boot` = median(delta_sd / boot_sd),
            q10 = quantile(delta_sd/boot_sd, .1),
            q90 = quantile(delta_sd/boot_sd, .9),
            `min delta SD` = min(delta_sd), .groups = "drop")
knitr::kable(summ_n, digits = 3,
             caption = "Mode axis: per-quantity agreement. Ratio near 1 across all 11 quantities.")
```

```{r scatter-mode, fig.height=6.6, fig.cap="Mode axis: per-mode delta SD vs bootstrap SD."}
ggplot(datn, aes(boot_sd, delta_sd, colour = group)) +
  geom_abline(slope = 1, colour = "grey60") +
  geom_point(alpha = .4, size = .5) +
  facet_wrap(~quantity, scales = "free", ncol = 3) +
  scale_colour_manual(values = c("phi decomposition" = "#c1442e", profiles = "#2e6fc1")) +
  labs(x = "bootstrap SD (uniform-resample)", y = "delta SD", colour = NULL) +
  theme_minimal(9) + theme(legend.position = "top")
```

```{r ratio-mode, fig.height=6.6, fig.cap="Mode axis: per-mode ratio delta/bootstrap (1.0 = agreement)."}
ggplot(datn, aes(mode, delta_sd / boot_sd, colour = group)) +
  geom_hline(yintercept = 1, colour = "grey60") +
  geom_point(alpha = .4, size = .5) +
  facet_wrap(~quantity, scales = "free_y", ncol = 3) +
  scale_colour_manual(values = c("phi decomposition" = "#c1442e", profiles = "#2e6fc1")) +
  coord_cartesian(ylim = c(0.7, 1.3)) +
  labs(x = "mode", y = "delta / bootstrap", colour = NULL) +
  theme_minimal(9) + theme(legend.position = "top")
```

\clearpage

# Per-quantity delta $\leftrightarrow$ bootstrap correspondence

```{r corr-table}
corr <- tribble(
  ~Quantity, ~`Delta (shipped)`, ~`Bootstrap`,
  "MM/MS/MA/MSA (uncentred)", "sqrt(sum_k h^2), h=spm_hmat(M,w,FALSE)", "SD of 0.5*log(Hajek msd_v), uniform draw",
  "MM/MS/MA/MSA (centred)",   "sqrt(sum_k h^2), h=spm_hmat(M,w,TRUE)",  "SD of (0.5*log msd_v - mean), same draw",
  "phi_mut",  "= nlrmsd_mm band (single MM model)",                   "SD of nlrmsd_mm draws",
  "phi_stab", "sqrt(sum_k (h_ms - h_mm)^2), shared scan",             "SD of (nlrmsd_ms - nlrmsd_mm), shared draw",
  "phi_act",  "sqrt(sum_k (h_msa - h_ms)^2), shared scan",            "SD of (nlrmsd_msa - nlrmsd_ms), shared draw")
knitr::kable(corr, caption = "What each quantity's delta formula and bootstrap procedure are.")
```

# Conclusion

On **both** the site axis and the mode axis, all eleven quantities — the four nested variants in
uncentred and centred form, and the three decomposition contributions — agree between the
shipped delta formula and the independent bootstrap to within Monte-Carlo noise. The delta
SPM band is validated.
